---
title: "Correlation"
subtitle: "Correlation Visualisation & Measures"
author: "Francisco Rowe"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html:
      number_sections: true
      css: extra.css
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library(knitr)
library(tidyverse)
library(kableExtra)
library(Hmisc) # compute statistical significance of correlations
library(ggcorrplot) # visualise correlation matrices
library(RColorBrewer) # nice colour schemes
library(viridis) # nice colour schemes
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```


In this session, we turn our focus on understanding how we can measure the relationship between two variables. 

We are often interested in associations eg. How is unemployment associated with education? How is commuting associated with income?

# Read data

```{r}
# clean workspace
rm(list=ls())
# load data
load("../data/data_census.RData")
```


```{r, eval=FALSE}
#introducing the data
head(census)
```

* Each row corresponds to a district

* The dataset records the % of persons in each district falling into a range of socio-demographic categories, except for the first five columns, which record geographic information.

**TASK #1** Explore the structure of the data

# Correlation: The Logic



```{r, fig.margin = TRUE, fig.cap = 'Fig.1 relationship unemployment and education across UK districts'}
ggplot(data=census) +
  geom_point( aes(y= Unemployed, x= No_Quals) )   +
  geom_smooth(aes(y= Unemployed, x= No_Quals), method = "lm", se=FALSE) +
    # Add labels
    labs(title= paste(" "), y="Unemployed (%)", x="No Qualification (%)") +
    theme_classic() +
    theme(axis.text=element_text(size=14))
```

*Correlation coefficient* measures the strength of the relationship between two variables.

## Correlation & Type of Data

Measure | Type of Data | Range
----------|---------------------------------------------
Pearson | symmetrical continuous distributions | -1 to +1
Spearman Rank | one or both skewed distributions | -1 to +1
Spearman Rank | both ordinal | -1 to +1
Cramer's V | one or both nominal | 0 to +1

In practice: Cramer's V is rarely used.

# Correlation

*Correlation between continuous variables*

```{r}
attach(census)
# Pearson correlation
cor( No_Quals, Unemployed, method="pearson")
# Spearman correlation
cor( No_Quals, Unemployed, method="spearman")
```

Between all possible combinations of variables in a data frame:

```{r, eval=FALSE}
pc <- cor( census[ , -c(1:5) ], method="pearson" )
round(pc, 2)
```


**TASK #2** Identify the 3 variables most strongly and most weakly correlated with the % of residents in ill health (illness).

**TASK #3** Create graphs visualising examples of strong, moderate and weak correlations.
```{r, include=FALSE}
# eg.
ggplot(data=census) +
  geom_point( aes(y= illness, x= Professionals) )   +
    # Add labels
    labs(title= paste(" "), y="Illness (%)", x="Professionals (%)") +
    theme_classic() +
    theme(axis.text=element_text(size=14))
```

# Testing Statistical Significance

We use a different function `rcorr` from the `Hmisc` package.
```{r}
pc <- rcorr(as.matrix(census[, 6:10]), type = "pearson")
pc
```

# Visualisation

Visualising correlation matrices

```{r, fig.margin = TRUE, fig.cap = 'Fig.2 Correlogram'}
# get correlations
pc <- cor( census[ , -c(1:5) ], method="pearson" )
ggcorrplot(pc)
```
You can adjust the options and add the statistical significance:

```{r}
# get p-values
sig <- cor_pmat(pc)
# draw correlogram
ggcorrplot(pc, method = "square", type= "upper", 
          ggtheme = ggplot2::theme_classic,
          hc.order= TRUE, colors = brewer.pal(n = 3, name = "RdBu"), 
          outline.col = "white", lab = FALSE,
          p.mat = sig) + scale_fill_viridis(option="inferno")
```

There are more functions and packages you can explore: 

* `corrplot` - corrplot package

* `ggcor` - ggally package

*Correlation between continuous and categorical variables*

Using the QLFS and density distributions:

```{r, echo=FALSE}
rm(list=ls())
# load data
load("../data/data_qlfs.RData")
attach(qlfs)
# create new data frame
df <- qlfs %>% filter(!is.na(NetPay)) %>% 
  filter(!is.na(EthnicGroup)) %>% 
  filter(NetPay >= 0)
# remember to:
detach(qlfs)
# and:
attach(df)
```


```{r, fig.margin = TRUE, fig.cap = 'Fig.3 Net Pay Density by Ethicity'}
ggplot(data= df) +
  geom_density(alpha=0.5, colour="#FF6666", aes(x = NetPay, fill = EthnicGroup))
```

Another way to visualise data distributions is using boxplots:

```{r, warning=FALSE}
ggplot(data = df) +
  geom_boxplot(aes(x = EthnicGroup, y= NetPay, fill= EthnicGroup)) +
  theme(axis.text.x  = element_text(angle=90, vjust=0.5, size=8), legend.position="none") +
  scale_x_discrete(name="Ethnic Group") +
  scale_y_continuous(limits = c(0, 1500), name="Weight (Net Pay (weekly))")
  
```

# Appendix: Concepts and Functions to Remember

Function | Description
----------|---------------------------------------------
cor() | compute Pearson's (method="pearson") or Spearman's (method="spearman") correlation
rcorr(), cor_pmat() | compute statistical significance
